---
title: "How Rich Are the Continents? GDP per Capita, 1952 vs 2007 (R)"
author: "Rachel Yan"
date: "2026-09-25"
---

In this post I use R to compare average GDP per capita across continents
in 1952 and 2007, and to see which continents grew the fastest.

Data: [gapminder](https://cran.r-project.org/package=gapminder) R package by
Jennifer Bryan, an excerpt of data from the
[Gapminder Foundation](https://www.gapminder.org/data/) (CC BY 4.0).
GDP per capita is measured in inflation-adjusted US dollars.

## Setup

```{r}
#| message: false
#| warning: false
library(gapminder)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Explore the data

```{r}
dim(gapminder)
range(gapminder$year)
gapminder |> count(continent, name = "rows")
```

## Average GDP per capita by continent

```{r}
gdp_by_year <- gapminder |>
  filter(year %in% c(1952, 2007)) |>
  group_by(continent, year) |>
  summarise(mean_gdp = round(mean(gdpPercap), 0), .groups = "drop")

gdp_by_year |>
  pivot_wider(names_from = year, values_from = mean_gdp) |>
  mutate(times = round(`2007` / `1952`, 1)) |>
  arrange(desc(`2007`))
```

## The richest countries in 1952

```{r}
gapminder |>
  filter(year == 1952) |>
  arrange(desc(gdpPercap)) |>
  select(country, continent, gdpPercap) |>
  head(5)
```

## Visualise

```{r}
#| label: fig-gapminder-r
#| fig-cap: "Average GDP per capita by continent in 1952 and 2007."
ggplot(gdp_by_year, aes(x = continent, y = mean_gdp, fill = continent)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::dollar(mean_gdp)), vjust = -0.4, size = 3) +
  facet_wrap(~ year) +
  scale_y_continuous(labels = scales::label_dollar(),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("#F7A6AC", "#EEC78A", "#CBE4B1", "#B8E5FA", "#EEE9A2")) +
  labs(x = NULL, y = "Average GDP per capita (US$)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## What I found

Average GDP per capita grew on every continent between 1952 and 2007, but the
gaps between continents became much larger. Oceania was the richest continent
in both years, rising from US$10,298 to US$29,810. Europe came second, reaching
US$25,054 in 2007. Africa remained by far the poorest: its average rose from
US$1,253 to only US$3,089, about a tenth of Oceania's in 2007.

Europe grew the fastest, with average GDP per capita rising 4.4 times. The other
continents grew between 2.4 and 2.9 times. Because Africa started from the
lowest level and grew relatively slowly, the absolute gap between Africa and
Oceania widened from about US$9,000 in 1952 to almost US$27,000 in 2007.

Averages can be misleading. In 1952, Kuwait had a GDP per capita of about
US$108,000 thanks to its oil wealth, more than seven times that of the next
richest country, Switzerland (about US$14,700). This single country pulls up
Asia's 1952 average and makes Asia look richer than the Americas, even though
most Asian countries were poor at the time. It also helps explain why Asia shows
the smallest growth ratio (2.4 times), despite the rapid economic growth of many
Asian countries. A median, or looking at individual countries, would give a
fuller picture. The data also stop in 2007.